---
title: "Lab 05 - Wrangling Spatial Data"
author: "Anahatt Virk"
date: "06/28/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

```{r dennys-alaska}
dn_ak <- dennys %>%
  filter(state == "AK")
  nrow(dn_ak)
```

There are `r nrow(dn_ak)` Denny's locations in Alaska.

```{r laquinta-alaska}
lq_ak <- laquinta %>%
  filter(state == "AK")
  nrow(lq_ak)
```

There are `r nrow(lq_ak)` La Quinta hotels in Alaska.

### Exercise 2

```{r number-of-pairings}
num_dennys <- nrow(dn_ak)
num_lq <- nrow(lq_ak)

num_pairings <- num_dennys * num_lq
```

There are `r num_pairings` distances that we need to calculate.

### Exercise 3

```{r joining-dataframes}
dn_lq_ak <- full_join(dn_ak, lq_ak, by = "state")

dn_lq_ak
```

### Exercise 4

In the joined data frame there are 6 total observations. The variables include address, city, state, zip code, longitude, and latitude. Each of these variables repeats twice as .x and .y, one coming from the Denny's dataset, and the other coming from the La Quinta dataset. However, the state variable only appears once as it was the join key. 

### Exercise 5

```{r add-variable}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

R <- 6371 # Earth mean radius in km

a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
d <- R * 2 * asin(sqrt(a))

return(round(d, round)) # distance in km
}
```

### Exercise 6

```{r calculating-distance}
dn_lq_ak <- dn_lq_ak %>%
  mutate(distance = haversine(
    long1 = longitude.x,
    lat1 = latitude.x,
    long2 = longitude.y,
    lat2 = latitude.y,
    round = 1))
```

### Exercise 7

```{r mindist}
dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_ak_mindist
```

### Exercise 8

```{r summarystats}
summary(dn_lq_ak_mindist)
```

```{r visualization}
ggplot(dn_lq_ak_mindist, aes(y = closest)) +
  geom_boxplot() + 
  labs(title = "Distances from Denny's to Nearest La Quinta (Alaska)", y = "Distance (km)") + 
  theme_minimal()

ggsave("dennys_la_quinta_distances.png")
```

The distances from Denny's to their nearest La Quinta locations in Alaska range from 2 km to 6 km, with an average of 4.4 km. The median distance is 5.2 km. 

### Exercise 9

```{r dennys-nc}
dn_nc <- dennys %>%
  filter(state == "NC")
  nrow(dn_nc)
```

There are `r nrow(dn_nc)` Denny's locations in North Carolina. 

```{r laquinta-nc}
lq_nc <- laquinta %>%
  filter(state == "NC")
  nrow(lq_nc)
```

There are `r nrow(lq_nc)` La Quinta hotels in North Carolina. 

```{r nc-pairings}
num_dennys_nc <- nrow(dn_nc)
num_lq_nc <- nrow(lq_nc)

num_pairings_nc <- num_dennys_nc * num_lq_nc
```

There are `r num_pairings_nc` distances that we need to calculate.

```{r nc-joining}
dn_lq_nc <- full_join(dn_nc, lq_nc, by = "state")

dn_lq_nc
```

```{r nc-add-variable}
dn_lq_nc <- dn_lq_nc %>%
  mutate(distance = haversine(
    long1 = longitude.x,
    lat1 = latitude.x,
    long2 = longitude.y,
    lat2 = latitude.y,
    round = 1))
```

```{r nc-mindist}
dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_nc_mindist
```

```{r summarystats-nc}
summary(dn_lq_nc_mindist)
```

```{r visualization-nc}
ggplot(dn_lq_nc_mindist, aes(y = closest)) +
  geom_boxplot() + 
  labs(title = "Distances from Denny's to Nearest La Quinta (North Carolina)", y = "Distance (km)")+
  theme_minimal()

ggsave("dennys_nc_distances.png")
```

The distances from Denny's to their nearest La Quinta locations in North Carolina range from 1.8 km to 187.9 km, with an average of 65.44 km. The median distance is 53.45 km. 

### Exercise 10

```{r dennys-tx}
dn_tx <- dennys %>%
  filter(state == "TX")
  nrow(dn_tx)
```

There are `r nrow(dn_tx)` Denny's locations in Texas. 

```{r laquinta-tx}
lq_tx <- laquinta %>%
  filter(state == "TX")
  nrow(lq_tx)
```

There are `r nrow(lq_tx)` La Quinta hotels in Texas.

```{r tx-pairings}
num_dennys_tx <- nrow(dn_tx)
num_lq_tx <- nrow(lq_tx)

num_pairings_tx <- num_dennys_tx * num_lq_tx
```

There are `r num_pairings_tx` distances that we need to calculate.

```{r tx-joining}
dn_lq_tx <- full_join(dn_tx, lq_tx, by = "state")

dn_lq_tx
```

```{r tx-add-variable}
dn_lq_tx <- dn_lq_tx %>%
  mutate(distance = haversine(
    long1 = longitude.x,
    lat1 = latitude.x,
    long2 = longitude.y,
    lat2 = latitude.y,
    round = 1))
```

```{r tx-mindist}
dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_tx_mindist
```

```{r summarystats-tx}
summary(dn_lq_tx_mindist)
```

```{r visualization-tx}
ggplot(dn_lq_tx_mindist, aes(y = closest)) +
  geom_boxplot() + 
  labs(title = "Distances from Denny's to Nearest La Quinta (Texas)", y = "Distance (km)")+
  theme_minimal()

ggsave("dennys_tx_distances.png")
```

The distances from Denny's to their nearest La Quinta locations in Texas range from 0.0 km to 60.6 km, with an average of 5.8 km. The median distance is 3.4 km. 

### Exercise 11

```{r dennys-nj}
dn_nj <- dennys %>%
  filter(state == "NJ")
  nrow(dn_nj)
```

There are `r nrow(dn_nj)` Denny's locations in New Jersey.

```{r laquinta-nj}
lq_nj <- laquinta %>%
  filter(state == "NJ")
  nrow(lq_nj)
```

There are `r nrow(lq_nj)` La Quinta hotels in New Jersey.

```{r nj-pairings}
num_dennys_nj <- nrow(dn_nj)
num_lq_nj <- nrow(lq_nj)

num_pairings_nj <- num_dennys_nj * num_lq_nj
```

There are `r num_pairings_nj` distances that we need to calculate.

```{r nj-joining}
dn_lq_nj <- full_join(dn_nj, lq_nj, by = "state", relationship = "many-to-many")

dn_lq_nj
```

```{r nj-add-variable}
dn_lq_nj <- dn_lq_nj %>%
  mutate(distance = haversine(
    long1 = longitude.x,
    lat1 = latitude.x,
    long2 = longitude.y,
    lat2 = latitude.y,
    round = 1))
```

```{r nj-mindist}
dn_lq_nj_mindist <- dn_lq_nj %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))

dn_lq_nj_mindist
```

```{r summarystats-nj}
summary(dn_lq_nj_mindist)
```

```{r visualization-nj}
ggplot(dn_lq_nj_mindist, aes(y = closest)) +
  geom_boxplot() + 
  labs(title = "Distances from Denny's to Nearest La Quinta (New Jersey)", y = "Distance (km)")+
  theme_minimal()

ggsave("dennys_n=j_distances.png")
```

The distances from Denny's to their nearest La Quinta locations in New Jersey range from 15.4 km to 69.1 km, with an average of 42.3 km. The median distance is 41.5 km. 

### Exercise 12

Among the states examined, Mitch Hedberg's joke is most likely to hold true in Texas. Texas has the highest number of both Denny's and La Quinta locations, and their locations are extremely close. The lowest distance between them is 0.0 km, and the mean distance is 5.8 km, reflecting their close proximity to one another in Texas. While the greatest distance is 60.6 km, the low median and mean suggest that this is likely an outlier.
