---
title: "Lab 05 - Wrangling Spatial Data"
author: "Anahatt Virk"
date: "06/28/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

```{r dennys-alaska}
dn_ak <- dennys %>%
  filter(state == "AK")
  nrow(dn_ak)
```

There are `r nrow(dn_ak)` Denny's locations in Alaska.

```{r laquinta-alaska}
lq_ak <- laquinta %>%
  filter(state == "AK")
  nrow(lq_ak)
```

There are `r nrow(lq_ak)` La Quinta hotels in Alaska.

### Exercise 2

```{r number-of-pairings}
num_dennys <- nrow(dn_ak)
num_lq <- nrow(lq_ak)

num_pairings <- num_dennys * num_lq
```

There are `r num_pairings` distances that we need to calculate.

### Exercise 3

```{r joining-dataframes}
dn_lq_ak <- full_join(dn_ak, lq_ak,
  by = "state")

dn_lq_ak
```

### Exercise 4

In the joined data frame there are 6 total observations. The variables include 

### Exercise 5

...

### Exercise 6

...

### Exercise 7

### Exercise 8

### Exercise 9

### Exercise 10

### Exercise 11

### Exercise 12
